Towards Semantic Patches

### 2010-06-18 papers/MIPS2010/abstract.tex
* Un peu de texte en plus.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@104 403e655e-980a-47bd-8c60-1ee5348bb6b7

\paragraph{Typing} The typing of our system is a restriction of the
usual typing rules for dependently-typed systems with the additional
two rules for typing a singleton product and the application of a
singleton argument (featuring the syntactic verification in the
environment). It depends on: 
\begin{itemize}
  \item$\mathcal S$ a set of sorts \sort,
  \item$\mathcal A \in \mathcal S^2$ the axioms of the system,
  \item$\mathcal R \in \mathcal S^3$ the allowed products.
\end{itemize}
It relies on two distincts judgments:
\begin{itemize}
\item $\Gamma\vdash a : t$ ($a$ has type $t$ in $\Gamma$),
\item $\Gamma\vdash t : \sort$ ($t$ has sort $s$ in $\Gamma$).
\end{itemize}
The rules are:
\begin{mathpar}
  \infer[Axiom]{
  }{\Gamma\vdash\sort_1:\sort_2}
  \quad(\sort_1,\sort_2)\in\mathcal A
  \and
  \infer[Init]{
    \Gamma(x):t
  }{\Gamma\vdash x:t}
  \and
  \infer[Prod]{\Gamma\vdash t:\sort_1 \and 
    \Gamma[x:t]\vdash u:\sort_2
  }{\Gamma\vdash(x:t)\cdot u : \sort_3}
  \quad(\sort_1,\sort_2,\sort_3)\in\mathcal R
  \and
  \infer[Sing]{\Gamma\vdash a:t \and
    \Gamma\vdash t:\sort_1 \and
    \Gamma[x=a:t]\vdash u:\sort_2
  }{\Gamma\vdash(x=a:t)\cdot u : \sort_3} 
  \quad(\sort_1,\sort_2,\sort_3)\in\mathcal R
  \and
  \infer[App]{
    \Gamma\vdash a:(y:t)\cdot u \and \Gamma(x):t
    }{\Gamma\vdash a\ x : u\subst{x}{y}}
  \and
  \infer[Skip]{
    \Gamma\vdash a:(y=b:t)\cdot u \and \Gamma(x)=b:t
    }{\Gamma\vdash a : u\subst{x}{y}}
\end{mathpar}

@@

The middle layer is a mechanized formalization of a metatheory. This
module provides a syntax for the objects of the metatheory (formulae,
judgments, \ldots) as well as the admissible rules (that we also call
metatheorems).  An \textit{internalization} function $\phi$ translates
the constructors of these objects and the specification of the
admissible rules as terms of the kernel language. As a consequence,
every object~$t$ of the theory can be injected into the
repository~$R$. An \textit{externalization} function $\psi$ is
implementing the inverse operation. Given a modified version $t'$ of
an object~$t$, we are able to compute the patch $\Delta$ using
the \textit{differienciation} function $\nabla(t)$.

@@

The implementation of this architecture raises some interesting
technical difficulties that can be handled, but only partially, 
by existing technology. Indeed, even if internalization and 
externalization are easily implemented thanks to the expressivity
of the kernel language, the projection, putback, differenciation
and integration operations are more intricate to realize.

@@

The problem of updating some data according to a modification that was
done on a partial view of it is a well-known issue in database
litterature called the \textit{view-update} problem. Pierce \textit{et
  al} have recently attacked it from a language design approach. Their
idea is to develop projection and putback functions simultaneously by
writing a so-called lens in a bidirectional programming
language~\cite{pierce-bidirectional}. Roughly speaking, a lense is a
software component that can be used to produce output from input but
also input from output (hence the bidirectionality). A lense is able
to keep track of the information that was erased by a projection to
produce a projected output from the initial input. When we supply a
modified output to the lense, it can elaborate, from the initial input
and under certain conditions, the missing part of the modified input.

We plan to design a bidirectional language to define the
projection~function~$\pi$ and the putback~function~$\sigma$ as a
lense. Yet, we cannot just import existing lenses combinators from the
Harmony system that already works on trees. Indeed, most of the time,
our trees have extra structural constraints induced by the presence of
binders.

@@

Computing the minimal edition script between two abstract syntax trees
is a NP-hard problem that has been heavily studied, in particular in
bioinformatics.  We guess that, in practice, we can restrict the
search to local transformations that preserve substructures (that
roughly consist in inserting, removing or moving around entire
subtrees). In that case, an existing polynomial algorithm might be
used~\cite{REF!}. Besides, abstract trees are relativily small in
practice.

We are more concerned about our ability to elaborate high-level patchs
from these low-level edition scripts. Actually, it is easy to deduce
from a low-level edition script, a low-level patch, based on object
syntax constructor elimination and introduction. However, such a
low-level patch may forbid a more efficient reuse of existing
objects. For instance, let us assume a transformation of an
object-level term~$x + x$ into~ he term~$2 + 2$. A low-level patch
would represents the transformation as the replacement of the first
subterm $x$ by the subterm~$2$ followed by the replacament of the
second subterm $x$ by the subterm~$2$. If we had an object-level type
derivation for $x + x$, the integration function would have to build
two subderivations, one for each replaced subterms.  Instead, a
high-level patch would represent that transformation as a substitution
of the free occurrences of~$x$ by the term $2$ in the term~$x + x$. In
that case, the integration function has the opportunity to apply a
metatheorem about the stability of typing derivations under
substitution and finally reuse the initial typing derivation
directly. Inferring automatically such a high-level patch seems an
unreachable objective~: there is clearly no most general high-level
patch. Nonetheless, an interactive process can be designed to help the
user building his high-level patch from a set of pre-computed
interpretations of the low-level patch.

@@

@@

\remtext{Algorithme d'élaboration de patchs?}

>>>>

\paragraph{Typing} The typing of our system is a restriction of the
usual typing rules for dependently-typed systems with the additional
two rules for typing a singleton product and the application of a
singleton argument (featuring the syntactic verification in the
environment). It depends on: 
\begin{itemize}
  \item$\mathcal S$ a set of sorts \sort,
  \item$\mathcal A \in \mathcal S^2$ the axioms of the system,
  \item$\mathcal R \in \mathcal S^3$ the allowed products.
\end{itemize}
It relies on two distinct judgments:
\begin{itemize}
\item $\Gamma\vdash a : t$ ($a$ has type $t$ in $\Gamma$),
\item $\Gamma\vdash t : \sort$ ($t$ has sort $s$ in $\Gamma$).
\end{itemize}
The rules are:
\begin{mathpar}
  \infer[Axiom]{
  }{\Gamma\vdash\sort_1:\sort_2}
  \quad(\sort_1,\sort_2)\in\mathcal A
  \and
  \infer[Init]{
    \Gamma(x):t
  }{\Gamma\vdash x:t}
  \and
  \infer[Prod]{\Gamma\vdash t:\sort_1 \and 
    \Gamma[x:t]\vdash u:\sort_2
  }{\Gamma\vdash(x:t)\cdot u : \sort_3}
  \quad(\sort_1,\sort_2,\sort_3)\in\mathcal R
  \and
  \infer[Sing]{\Gamma\vdash a:t \and
    \Gamma\vdash t:\sort_1 \and
    \Gamma[x=a:t]\vdash u:\sort_2
  }{\Gamma\vdash(x=a:t)\cdot u : \sort_3} 
  \quad(\sort_1,\sort_2,\sort_3)\in\mathcal R
  \and
  \infer[App]{
    \Gamma\vdash a:(y:t)\cdot u \and \Gamma(x):t
    }{\Gamma\vdash a\ x : u\subst{x}{y}}
  \and
  \infer[Skip]{
    \Gamma\vdash a:(y=b:t)\cdot u \and \Gamma(x)=b:t
    }{\Gamma\vdash a : u\subst{x}{y}}
\end{mathpar}

@@

The middle layer is a mechanized formalization of a metatheory. This
module provides a syntax for the objects of the metatheory (formulae,
judgments, \ldots) as well as the admissible rules (that we also call
metatheorems).  An \textit{internalization} function $\phi$ translates
the constructors of these objects and the specification of the
admissible rules as terms of the kernel language. As a consequence,
every object~$t$ of the theory can be injected into the
repository~$R$. An \textit{externalization} function $\psi$ is
implementing the inverse operation. Given a modified version $t'$ of
an object~$t$, we are able to compute the patch $\Delta$ using
the \textit{differentiation} function $\nabla(t)$.

@@

The implementation of this architecture raises some interesting
technical difficulties that can be handled, but only partially, 
by existing technology. Indeed, even if internalization and 
externalization are easily implemented thanks to the expressivity
of the kernel language, the projection, putback, differentiation
and integration operations are more intricate to realize.

@@

The problem of updating some data according to a modification that was
done on a partial view of it is a well-known issue in database
literature called the \textit{view-update} problem. Pierce \textit{et
  al} have recently attacked it from a language design approach. Their
idea is to develop projection and putback functions simultaneously by
writing a so-called lens in a bidirectional programming
language~\cite{pierce-bidirectional}. Roughly speaking, a lens is a
software component that can be used to produce output from input but
also input from output (hence the bidirectionality). A lens is able
to keep track of the information that was erased by a projection to
produce a projected output from the initial input. When we supply a
modified output to the lens, it can elaborate, from the initial input
and under certain conditions, the missing part of the modified input.

We plan to design a bidirectional language to define the
projection~function~$\pi$ and the putback~function~$\sigma$ as a
lens. Yet, we cannot just import existing lenses combinators from the
Harmony system that already works on trees. Indeed, most of the time,
our trees have extra structural constraints induced by the presence of
binders.

@@

Computing the minimal edition script between two abstract syntax trees
is a NP-hard problem that has been heavily studied, in particular in
bio-informatics.  We guess that, in practice, we can restrict the
search to local transformations that preserve substructures (that
roughly consist in inserting, removing or moving around entire
sub-trees). In that case, an existing polynomial algorithm might be
used~\cite{REF!}. Besides, abstract trees are relatively small in
practice.

We are more concerned about our ability to elaborate high-level patches
from these low-level edition scripts. Actually, it is easy to deduce
from a low-level edition script, a low-level patch, based on object
syntax constructor elimination and introduction. However, such a
low-level patch may forbid a more efficient reuse of existing
objects. For instance, let us assume a transformation of an
object-level term~$x + x$ into~ he term~$2 + 2$. A low-level patch
would represents the transformation as the replacement of the first
sub-term $x$ by the sub-term~$2$ followed by the replacement of the
second sub-term $x$ by the sub-term~$2$. If we had an object-level type
derivation for $x + x$, the integration function would have to build
two sub-derivations, one for each replaced sub-terms.  Instead, a
high-level patch would represent that transformation as a substitution
of the free occurrences of~$x$ by the term $2$ in the term~$x + x$. In
that case, the integration function has the opportunity to apply a
metatheorem about the stability of typing derivations under
substitution and finally reuse the initial typing derivation
directly. Inferring automatically such a high-level patch seems an
unreachable objective~: there is clearly no most general high-level
patch. Nonetheless, an interactive process can be designed to help the
user building his high-level patch from a set of pre-computed
interpretations of the low-level patch.

@@

There is several ways of integrating a patch $\Delta$ inside a
repository~$R$.  The naive attempt would be a simple concatenation
of~$\Delta$ at the end of~$R$.  This kind of integration, though
possible and correct, is of little interest because it does not take
into account the fact that $\Delta$ probably defines new versions of
objects already present in~$R$.

A more useful integration propagates changes to the context that uses
the modified object in order to trigger the updating this context
too. Again, in full generality, it is not possible to update the
context automatically. Instead, we choose to generate
\textit{challenges}, which are adaptation patches with holes. These
challenges can be addressed by the author that wrote the patch to be
integrated, by the author of the context or automatically when there
is no ambiguity (think of a context that uses a function whose
body has changed but not its type).

@@
